maze-dataset

### 2025-03-29 docs/paper/paper.tex
diagram in pdf working
wip paper

\href{https://understanding-search.github.io/maze-dataset/examples/maze_examples.html}{examples}
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html}{docs}

@@

\textbf{TODO: mention the examples, put in a figure
https://understanding-search.github.io/maze-dataset/examples/maze\_examples.html}

\begin{figure} 
\begin{tikzpicture}[remember picture]
  % Place a clickable blue box at the top left corner (offset down by 1cm)
  \node[draw=blue, fill=blue!20, fill opacity=0.3,
        hyperlink node url={https://github.com},
        minimum width=2cm, minimum height=1cm,
        anchor=north west]
    at ($(current page.north west)+(0,-1cm)$) {};
\end{tikzpicture}
\caption{testinggggggggggggggggg}
\end{figure}

\newpage

@@

\hypertarget{features-and-implementation}{%
\section{Features and
Implementation}\label{features-and-implementation}}

@@

\textbf{TODO: JOSS says don't document the API itself. we should talk
about the capabilities, but just rewrite this to link to the
documentation.}

@@

To create a dataset, we first create a \texttt{MazeDatasetConfig}
configuration object, which specifies the seed, number, and size of
mazes, as well as the generation algorithm and its corresponding
parameters. This object is passed to a \texttt{MazeDataset} class to
create a dataset. Crucially, this \texttt{MazeDataset} inherits from a
PyTorch (\protect\hyperlink{ref-pytorch}{Paszke et al., 2019})
\texttt{Dataset}, and can thus be easily incorporated into existing data
pre-processing and training pipelines, e.g., through the use of a
\texttt{Dataloader} class.

@@

When initializing mazes, further configuration options can be specified
through the \texttt{from\_config()} factory method as necessary. Options
include 1) whether to generate the dataset during runtime or load an
existing dataset, 2) if and how to parallelize generation, and 3) where
to store the generated dataset. Full documentation of configuration
options is available in our repository
(\protect\hyperlink{ref-maze-dataset-github}{Michael I. Ivanitskiy et
al., 2023a}). Available maze generation algorithms are static methods of
the \texttt{LatticeMazeGenerators} class and include the following:

@@

\begin{figure}
\centering
\includegraphics{figures/output-fmts.pdf}
\caption{Various output formats. Top row (left to right): ASCII diagram,
rasterized pixel grid, and advanced display.}
\end{figure}

@@

In previous work, maze tasks have been used with Recurrent Convolutional
Neural Network (RCNN) derived architectures
(\protect\hyperlink{ref-deepthinking}{Schwarzschild, Borgnia, Gupta,
Huang, et al., 2021}). To facilitate the use of our package in this
context, we replicate the format of
(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}) and provide the \texttt{RasterizedMazeDataset}
class which returns rasterized pairs of (input, target) mazes as shown
in Figure below.

\begin{figure}
\centering
\includegraphics[width=0.3\textwidth,height=\textheight]{figures/maze-raster-input-target.pdf}
\caption{Input is the rasterized maze without the path marked (left),
and provide as a target the maze with all but the correct path removed.
Configuration options exist to adjust whether endpoints are included and
if empty cells should be filled in.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{figures/outputs-tokens-colored.tex}
\caption{Example text output format with token regions highlighted.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{figures/TokenizerElement_structure.pdf}
\caption{Nested internal structure of \texttt{\_TokenizerElement}
objects inside a typical \texttt{MazeTokenizerModular} object.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.95\textwidth,height=\textheight]{figures/gridsize-vs-gentime.pdf}
\caption{Plots of maze generation time. Generation time scales
exponentially with maze size for all algorithms (left). Generation time
does not depend on the number of mazes being generated, and there is
minimal overhead to initializing the generation process for a small
dataset (right). Wilson's algorithm is notably less efficient than
others and has high variance. Note that for both plots, values are
averaged across all parameter sets for that algorithm, and
parallelization is disabled.}
\end{figure}

>>>>

@@

Solving mazes is a classic problem in computer science and artificial
intelligence, and humans have been constructing mazes for thousands of
years. Although finding the shortest path through a maze is a solved
problem, this very fact makes it an excellent testbed for studying how
machine learning algorithms solve problems and represent spatial
information. We introduce \texttt{maze-dataset}, a Python library for
generating, processing, and visualizing datasets of mazes. This library
supports a variety of maze generation algorithms providing both mazes
with loops and ``perfect'' mazes without them. These generation
algorithms can be configured with various parameters, and the resulting
mazes can be filtered to satisfy desired properties. Also provided are
tools for converting mazes to and from various formats suitable for a
variety of neural network architectures, such as rasterized images and
tokenized text sequences, as well as various visualization tools. As
well as providing a simple interface for generating, storing, and
loading these datasets, \texttt{maze-dataset} is extensively tested,
type hinted, benchmarked, and documented.

\begin{figure} 
  \begin{minipage}{5in}
    \input{docs/paper/diagram/diagram.tikz} 
  \end{minipage}
  \caption{Usage of maze-dataset. We create a `MazeDataset` from a `MazeDatasetConfig`. This contains `SolvedMaze` objects which can be converted to and from a variety of formats. Code in the image contains clickable links to \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html}{documentation}. A variety of generated examples can be viewed \href{https://understanding-search.github.io/maze-dataset/examples/maze_examples.html}{here}.}
  \label{fig:diagram}
\end{figure}

@@

\hypertarget{features}{%
\section{Features}\label{features}}

@@

@@

To create a dataset, we first create a \texttt{MazeDatasetConfig}
configuration object, which specifies the seed, number, and size of
mazes, as well as the generation algorithm and its corresponding
parameters. This object is passed to a \texttt{MazeDataset} class to
create a dataset. Crucially, this \texttt{MazeDataset} mimics the
interface of a PyTorch (\protect\hyperlink{ref-pytorch}{Paszke et al.,
2019}) \texttt{Dataset}, and can thus be easily incorporated into
existing data pre-processing and training pipelines, e.g., through the
use of a \texttt{Dataloader} class.

@@

When initializing mazes, further configuration options can be specified
through the \texttt{from\_config()} factory method as necessary. Options
allow for saving/loading existing datasets instead of regenerating, and
parallelization options for generation. Available maze generation
algorithms are static methods of the \texttt{LatticeMazeGenerators}
class and include the following:

@@

\begin{figure}
\hypertarget{fig:visualoutputs}{%
\centering
\includegraphics{docs/paper/figures/output-fmts.pdf}
\caption{Various output formats. Top row (left to right): ASCII diagram,
rasterized pixel grid, and advanced display.}\label{fig:visualoutputs}
}
\end{figure}

@@

In previous work, maze tasks have been used with Recurrent Convolutional
Neural Network (RCNN) derived architectures
(\protect\hyperlink{ref-deepthinking}{Schwarzschild, Borgnia, Gupta,
Huang, et al., 2021}). To facilitate the use of our package in this
context, we replicate the format of
(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}) and provide the \texttt{RasterizedMazeDataset}
class which returns rasterized pairs of (input, target) mazes as shown
in \autoref{fig:e2hraster} below.

\begin{figure}
\hypertarget{fig:e2hraster}{%
\centering
\includegraphics[width=0.3\textwidth,height=\textheight]{docs/paper/figures/maze-raster-input-target.pdf}
\caption{Input is the rasterized maze without the path marked (left),
and provide as a target the maze with all but the correct path removed.
Configuration options exist to adjust whether endpoints are included and
if empty cells should be filled in.}\label{fig:e2hraster}
}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{docs/paper/figures/outputs-tokens-colored.tex}
\caption{Example text output format with token regions highlighted.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{docs/paper/figures/TokenizerElement_structure.pdf}
\caption{Nested internal structure of \texttt{\_TokenizerElement}
objects inside a typical \texttt{MazeTokenizerModular} object.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.95\textwidth,height=\textheight]{docs/paper/figures/gridsize-vs-gentime.pdf}
\caption{Plots of maze generation time. Generation time scales
exponentially with maze size for all algorithms (left). Generation time
does not depend on the number of mazes being generated, and there is
minimal overhead to initializing the generation process for a small
dataset (right). Wilson's algorithm is notably less efficient than
others and has high variance. Note that for both plots, values are
averaged across all parameter sets for that algorithm, and
parallelization is disabled.}
\end{figure}
